**The statistical analysis of non-linear longitudinal data in biomedical research

### 2021-01-28 Manuscripts/Manuscript_AM_v1.Rmd
File needs updates from master branch

---
title: '**The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models**'
author:
- Ariel Mundo^[Department of Biomedical Engineering, University of Arkansas, Fayetteville]
- Timothy J. Muldoon^[Department of Biomedical Engineering, University of Arkansas,
  Fayetteville]
- John R. Tipton^[Department of Mathematical Sciences, University of Arkansas, Fayetteville]
subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
output:
  bookdown::pdf_document2:
    keep_tex: yes
    fig_caption: yes
    extra_dependencies: ["subfig","breqn"]
  bookdown::word_document2: default
  bookdown::html_document2: default
  link-citations: yes
  css: style.css
  '': default
csl: ieee.csl
bibliography: refs.bib
---

@@

A longitudinal study is defined as that which is designed to repeatedly measure a variable of interest in a group (or groups) of subjects. In biomedical research, this type of study is employed when the intention is to observe the evolution of the effect of a certain treatment across time, rather than analyzing it at a single time point (a cross-sectional study). Clinical studies of breast and neck cancer have used a longitudinal approach [@sio2016;@kamstra2015]. In the first case, weekly measurements of skin toxicities in  patients with radiation-induced dermatitis were taken for up to 8 weeks; whereas in the latter mouth opening was assessed at 6,12, 18, 24 and 36 months after radiotherapy (RT). Longitudinal studies have beenn also used to analyze tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001;@roth2017], and cell metabolism [@jones2018;@skala2010].

Traditionally, a “frequentist” approach is used in biomedical research to derive inferences from a longitudinal study. Such statistical view derives its name from the fact that it regards probability as a limiting frequency[@wagenmakers2008] and its application is based on a null hypothesis test using the _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA). This methodology makes two key assumptions regarding longitudinal data: constant correlation exists across same-subject measurements, and observations from each subject are obtained at all time points through the study (a condition also known as complete observations) [@gueorguieva2004;@schober2018].

However, constant correlation is frequently unjustified as its value tends to diminish between measures when the time interval between them increases [@ugrinowitsch2004]. 
In the case of the number of observations during the study, different situations can arise that prevent the collection of complete measurements for all subjects:  in a clinical trial voluntary withdrawal from one or multiple patients can occur, whereas attrition in animals due to injury or weight loss can occur in preclinical experiments, and it is even possible that unexpected complications with equipment or supplies arise, preventing the researcher from collecting measurements at a certain time point and therefore violating the _complete observations_ assumption of rm-ANOVA.

When incomplete observations occur, an rm-ANOVA requires the exclusion of all subjects with missing observations from the analysis. This can result in increased costs if the desired statistical power is not met with the remaining observations, as it would be necessary to enroll more subjects. At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups. Additionally, rm-ANOVA uses a _post hoc_ analysis to assess differences between the measured response in different groups. A _post hoc_ analysis is based on multiple repeated comparisons to estimate a _p-value_, a metric that is widely used as a measure of significance. Because the _p-value_ is highly variable, multiple comparisons can inflate the false positivity rate [@liu2010;@halsey2015], consequently biasing the conclusions of the study.

During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal information. This is exemplified by the  use of  linear mixed effects models (LMEMs)  in certain groups to analyze longitudinal data [@skala2010;@vishwanath2009]. Briefly, these models incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g. the different drug regimens in a clinical trial), and _random effects_, which account for random variation within the population. These models are more flexible than the traditional rm-ANOVA as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject [@pinheiro2006]. On the other hand, they impose restrictions in the distribution of the errors  of the random effects [@gueorguieva2004].

One final assumption that is not initially evident for  both rm-ANOVA and LMEMs models pertains the relationship between the observed effect and the progression of the timeline of the study, which is expected to be linear [@pinheiro2006]. This common assumption in both rm-ANOVA ane LMEMs consequently restricts the inferences from the model when the data does not follow a linear trend. In biomedical research, a special case of this behavior in longitudinal data arises in measurements of tumor response in preclinical and clinical settings [@roblyer2011;@skala2010;@vishwanath2009]. These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of and the estimations of an LMEM or rm-ANOVA model inconsistent with the observed variation. In other words, the in both cases the model tries to explain highly-variable data using a linear trend, consequently biasing the estimates. Additionally,  although it is possible that a _post hoc_ analysis is able to find “significant” _p-values_( _p_<0.05) by using multiple comparisons between the model terms, the use of this estimator is limited in such scenario because it is intended to work in models that have a reasonable agreement with the data.

As  the “frequentist” rm-ANOVA and the more advanced LMEM approach are both limited in the analysis of non-linear longitudinal information, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the information (and not an assumed trend) to determine the fit of the model, while enabling inferences that are both adequate and consistent from a statistical perspective.
In this regard, Generalized Additive Models (GAMs) present an alternative approach to analyze longitudinal data. Although not commonly used in the biomedical community, these non-parametric models have been used to analyze temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018], health-environment interactions [@yang2012] and political science [@beck1998] . There are several advantages of GAMs over LMEMs and rm-ANOVA models: They have an extensively developed theoretical background, enable the data to dictate the trend of the model, can accommodate missing observations and do not require constant correlation between repeated measurements [@wood2017].Therefore, GAMs can provide a more suitable statistical method to analyze non-linear biomedical longitudinal data.

 The current advances in programming languages designed for statistical analysis (specifically $\textsf{R}$), have eased the computational implementation of more complex models beyond LMEMs. In particular, $\textsf{R}$ has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2011;@wood2016;@wood2017] that not only speed up the initial stages of the analysis but enable the use of advanced modeling structures (e.g. hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user. This programming language is also able to simulate data, an emerging strategy used to test statistical models [@haverkamp2017]. This allows to create and  explore different alternatives for analysis without collecting information in the field, and reduces the time window between experiment design and its implementation.

Therefore, the purpose of this study is to provide biomedical researchers with a clear understanding of the theory and the practical implementation of GAMs to analyze longitudinal data using by focusing on four areas. First, the limitations of rm-ANOVA and LMEMs to analyze longitudinal data are contrasted in the cases of missing observations, assumption of linearity and correlation structures. Secondly, the key theoretical elements of GAMS are presented without using extensive derivations or complex mathematical notation that can obscure their understanding.  Thirdly, simulated data that follows the trend of previously reported values [@vishwanath2009] is used to illustrate the type of non-linear longitudinal data that occurs in biomedical research and the implementation of GAMs in such scenario. Finally, reproducibility is emphasized by providing the code to generate the data and the implementation of GAMs in $\textsf{R}$, in conjunction with a step-by-step guide to fit models of increasing complexity.  
In summary, the exploration of modern statistical techniques to analyze longitudinal data may allow biomedical researchers to consider the use of GAMs instead of rm-ANOVA or LMEMs when the data does not follow a linear trend, and will also help to improve the standards for reproducibility  in biomedical research.

>>>>

---
title: '**The statistical analysis of non-linear longitudinal data in biomedical research
  using generalized additive models**'
author:
- Ariel Mundo^[Department of Biomedical Engineering, University of Arkansas, Fayetteville]
- Timothy J. Muldoon^[Department of Biomedical Engineering, University of Arkansas,
  Fayetteville]
- John R. Tipton^[Department of Mathematical Sciences, University of Arkansas, Fayetteville]
csl: ieee.csl
output:
- default
- default
- yes
- style.css
- default
- extra_dependencies:
  - subfig
  - breqn
  fig_caption: yes
  keep_tex: yes
subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
bibliography: refs.bib
---

@@

A longitudinal study is designed to repeatedly measure a variable of interest in a group (or groups) of subjects, when the intention is to observe the evolution of the effect of a certain treatment across time rather than analyzing it at a single time point (e.g., a cross-sectional study). Medical research commonly uses longitudinal studies to analyze the evolution of patients before, during and after treatment. For example, Sio et al analyzed the evolution of dermatitis symptoms in breast cancer patients after radiotherapy (RT), by taking weekly measurements over the course of two moths [@sio2016], and Kamstra et al measured mouth opening at regular monthly intervals after RT in neck cancer patients [@kamstra2015]. Biomedical research has used longitudinal studies as well to analyze tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001;@roth2017], and cell metabolism [@jones2018;@skala2010].

Traditionally, a “frequentist” or "classical" statistical paradigm is used in biomedical research to derive inferences from a longitudinal study. The frequentist paradigm regards probability as a limiting frequency[@wagenmakers2008] by assuming a null hypothesis using the _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA). This methodology makes two key assumptions regarding longitudinal data: constant correlation exists across same-subject measurements, and observations from each subject are obtained at all time points through the study (a condition also known as complete observations) [@gueorguieva2004;@schober2018].

The assumption of constant correlation is frequently unjustified as correlation in time tends to diminish between measurements when the time interval between them increases [@ugrinowitsch2004]. 
Due to a variety of causes, the number of observations during a study can vary between all subjects. For example, in a clinical trial voluntary withdrawal from one or multiple patients can occur, whereas attrition in animals due to injury or weight loss can occur in preclinical experiments, and it is even possible that unexpected complications with equipment or supplies arise, preventing the researcher from collecting measurements at a certain time point and therefore violating the _complete observations_ assumption of  classical rm-ANOVA.

When incomplete observations occur, an rm-ANOVA is fit by excluding all subjects with missing observations from the analysis[@gueorguieva2004]. This can result in increased costs if the desired statistical power is not met with the remaining observations, as it would be necessary to enroll more subjects. At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups. Additionally, rm-ANOVA uses a _post hoc_ analysis to assess differences between the measured response in different groups. A _post hoc_ analysis is based on multiple repeated comparisons to estimate a _p-value_, a metric that is widely used as a measure of significance. Because the _p-value_ is highly variable, multiple comparisons can inflate the false positivity rate [@liu2010;@halsey2015], consequently biasing the conclusions of the study.

During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal information. The recognition on the limitations of rm-ANOVA is exemplified by the  use of  linear mixed effects models (LMEMs) by certain groups to analyze longitudinal data [@skala2010;@vishwanath2009]. Briefly, LMEMs incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g. the different drug regimens in a clinical trial), and _random effects_, which account for random variation within the population. When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject [@pinheiro2006]. On the other hand, LMEMs impose restrictions in the distribution of the errors  of the random effects [@gueorguieva2004].

One final assumption that is not initially evident for both rm-ANOVA and LMEMs models is that the mean response is expected to change linearly through time[@pinheiro2006]. The linearity assumption in both rm-ANOVA ane LMEMs consequently restricts the inferences from the model when the data does not follow a linear trend. In biomedical research, a particular case of this non-linear behavior in longitudinal data arises in measurements of tumor response in preclinical and clinical settings [@roblyer2011;@skala2010;@vishwanath2009]. These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of and the estimations of an LMEM or rm-ANOVA model inconsistent with the observed variation. In other words, both LMEMs and rm-ANOVA explain highly-variable data using a linear trend, consequently biasing the estimates when the true response is non-linear in time.

Additionally,  although it is possible that a _post hoc_ analysis is able to find “significant” _p-values_( _p_<0.05) by using multiple comparisons between the model terms, the use of this estimator is limited in such scenario because it is intended to work in models that have a reasonable agreement with the data.

***
Additionally,  although it is possible that a _post hoc_ analysis is able to find “significant” _p-values_( _p_<0.05) it has to be reminded that the validity of such metric relies on how adequate the model is to the data. In other words, the _p-value_ requires that the model and the data have good agreement and if that is not the case, a "Type 3" error (known as "model misspecification") occurs[@dennis2019].That would be the case if for example, a model that is only able to explain linear responses is fitted to data that has a quadratic behavior. In such scenario, the _p-value_ becomes invalid[@wang2019].

As  the “frequentist” rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the information (and not an assumed trend) to determine the fit of the model, enabling inference that is appropiate to the data.
In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data. Although not commonly used in the biomedical community, these non-parametric models have been used to analyze temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018], health-environment interactions [@yang2012] and the dynamics of government in political science [@beck1998] . There are several advantages of GAMs over LMEMs and rm-ANOVA models: GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend of the model, can easily accommodate missing observations and can model non-constant correlation between repeated measurements [@wood2017].Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data.

 The current advances in programming languages designed for statistical analysis (specifically $\textsf{R}$), have eased the computational implementation of more complex models beyond LMEMs. In particular, $\textsf{R}$ has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016;@wood2017] that not only speed up the initial stages of the analysis but enable the use of advanced modeling structures (e.g. hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user. At the same time, $\textsf{R}$ has many tools that simplify data simulation, an emerging strategy used to test statistical models [@haverkamp2017]. This allows to create and  explore different alternatives for analysis without collecting information in the field, and reduces the time window between experiment design and its implementation.

The purpose of this work is to provide biomedical researchers with a clear understanding of the theory and the practical implementation of GAMs to analyze longitudinal data using by focusing on four areas. First, the limitations of rm-ANOVA and LMEMs to analyze longitudinal data are contrasted in the cases of missing observations, assumption of linearity of response and constant correlation structures. Second, the key theoretical elements of GAMs are presented without using extensive derivations or complex mathematical notation that can obscure their understanding.  Third, simulated data that follows the trend of previously reported values [@vishwanath2009] is used to illustrate the type of non-linear longitudinal data that occurs in biomedical research to highlight the differences in inference between rm-ANOVA, LMEMs and GAMs and the implementation of the latter in such scenario. Finally, reproducibility is emphasized by providing the code to generate the data and the implementation of GAMs in $\textsf{R}$, in conjunction with a step-by-step guide to fit models of increasing complexity.  In summary, the exploration of modern statistical techniques to analyze longitudinal data may allow biomedical researchers to consider the use of GAMs instead of rm-ANOVA or LMEMs when the data does not follow a linear trend, and will also help to improve the standards for reproducibility  in biomedical research.
